\assignment

### 2017-04-04 project/paper.rnw
Spell check.

One little-studied application of inhomogeneous point process intensity
estimation is the use of an intensity surface estimated from events in a
subregion to predict the intensity over the entire region of interest. This
procedure would be relevant whenver it is known or suspected that some type of
plant, animal, or other item occurs in the region, and the goal of the
analysis is to map the trend in where these tend to be located rather than
estimate parameters of some process at work across hypothetical replicates of
similar regions. It may be prohibitively expensive or difficult to observe all
events over the entire region, so a sample of subregions is taken. For
example, conservationists want to study the spatial distribution of an
endangered plant across a large region of thick jungle so that they can
establish a preserve where the plant is protected. They cannot search the
entire jungle, so they take a simple random sample of quadrats and record the
locations of the plants in those quadrats. They will then want to fit a model
describing the trends in intensity of these plants extrapolated over the
entire jungle. In this setting, the objective is to map the realized spatial
intensity of the plant over this jungle so that these plants can be protected,
not to estimate parameters of the process that arranges plants of this species
at jungles like this one.

Another situation where it is useful to extrapolate point pattern intensity
outside of the observed region is in mapping subsurface geomagnetic anomalies,
such as the munitions debris found at former military test ranges. This is
frequently done in the early stages of an unexploded ordnance (UXO)
remediation, where the intensity surface of inert munitions fragments is
used to identify the locations of targets so that the search for UXO can be
focused on the sections of the site most likely to contain it. The anomalies
are only observed when detection equipment passes directly over them, but the
project leaders are concerened with finding UXO and do not want to waste
resources finding every inert fragment that could be found by metal detectors.
The most common data collection method is to take a systematic sample of
straight-line transects and observe anomalies in rectangular regions centered
along the transects. Frequently, moving averages of the intensity are computed
in circular windows and an intensity map is produced using ordinary kriging to
predict the intensity in a grid of these windows. There are several problems
with this approach: it assumes stationarity when the moving averages are
believed to be non-stationary, it ignores the point process nature of the
data, and it is very sensitive to the window size~\parencite{vspguide,wp}. In
this project, I review the use of polynomial trend surface models for the
log-intensity at a simulated UXO site, ultimately concluding that these
methods are not yet developed well enough to be useful.

@@

The log-likelihood of an unmarked Poisson process on a region \(D\) with
intensity \(\lambda(\mathbf{s})\) is found by conditioning on the number of
events in the following manner. The number of events in \(D\) follows a
Poisson distribution with mean \(\mu=\int_{D}\lambda(\mathbf{s})d\mathbf{s}\).
Given that \(n\) events occured, their locations \(\mathbf{s}_{1},\dots,
\mathbf{s}_{n}\) are independent and identically distributed with density
\(\lambda(\mathbf{s})/\mu\). Then the log-likelihood of the intensity is
\begin{align*}
\ell(\lambda) &= \left\{-\mu + n\log(\mu) - \log(n!)\right\}
+ \sum_{i=1}^{n}\left\{\log\left(\lambda(\mathbf{s}_{i})\right)
- \log(\mu)\right\} \\
&= \sum_{i=1}^{n}\log\left(\lambda(\mathbf{s}_{i})\right)
-\int_{D}\lambda(\mathbf{s})d\mathbf{s} - \log(n!).
\end{align*}
A natural way to model the intensity is to define a log-linear model
\begin{equation*}
\log(\lambda(\mathbf{s})) = \mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
\end{equation*}
where \(\mathbf{x}(\mathbf{s})\) can, in principle, include functions of the
spatial coordinates and also covariates. Unfortunately, the likelihood becomes
\begin{equation*}
\sum_{i=1}^{n}\mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
-\int_{D}\exp\left(\mathbf{x}(\mathbf{s})^{T}
\boldsymbol{\beta}\right)d\mathbf{s} - \log(n!)
\end{equation*}
so \(\mathbf{x}(\mathbf{s})\) must be known (or predicted) either across the
entire region \(D\) or at enough locations to approximate the integral
numerically. Thus, covariates require some extra data collection or modeling
effort to incorporate, but trend surfaces are feasible.

\textcite{bermanturner} proposed a practical method for fitting these models
with generalized linear model (GLM) software using quadratic approximations to
the likelihood, even generalizing to other link functions besides the log link.
Their method is based on partitioning \(D\) such that each subset in the
partition contains at most one event, and then maximizing a weighted
pseudolikelihood based upon a binomial or Poission distribution for the count
of events in each subset. They comment that the covariance matrices produced
by standard GLM software under this method are reasonable approximations to
the correct covariance matrices that would be derived from the true likelihood
as long as the quadrature approximation of the likelihood converges to the
true likelihood as the resolution of the partition is increased.
This was followed up by \textcite{baddeleycoeurjolly14}, who developed a
logistic regression-based approach that avoids the quadratic approximation,
thereby reducing bias in the coefficient estimates. Both methods are available
in the \texttt{spatstat} package~\parencite{spatstat} for R~\parencite{baser}.
The \texttt{spatstat} function \texttt{ppm} implements these methods using the
base R \texttt{glm} function to maximize the likelihood; I will use the Berman
and Turner method because it is the default in \texttt{ppm} and therefore
better reflects the experience of a na\"{i}ve practitioner who is not an
expert on spatial point processes.

@@

In my writing project, I evaluated ordinary kriging of moving average
intensity values on a simulated tank training range. I will use the same site
for this project. The site is a
\Sexpr{sprintf('%.2f', area(site_window) / 43560)} acre quadrilateral region
with homogeneous Poisson process producing background anomalies with an
intensity of 100 anomalies per acre, and two regions of concentrated munitions
use generated as Poisson processes with bivariate Gaussian intensity. The
total intesnsity at the center of each concentrated munitions use region is
300 anomalies per acre~(Figure~\ref{fig:truth}).

<<truth, cache = TRUE, dependson = 'uxoexampledata', fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
# Bivariate normal intensity function that I originalally wrote for my
# writing project, parameterized in terms of the major and minor axes scales
# and the angle of the major axis from horizontal.
#   a is horizontal axis, b is vertical axis, r is rotation angle
gauss.elliptic <- function(x, y, mu.x = 0, mu.y = 0, s.a = 1, s.b = 1,
                           r = 0, maxrate = 1){
  rot <- zapsmall(matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2))
  ab <- diag(c(s.a^2, s.b^2))
  sigma <- rot %*% ab %*% t(rot)
  siginv <- solve(sigma)
  mu <- matrix(c(mu.x, mu.y), nrow = 2)
  mat <- matrix(rbind(x, y), nrow = 2)
  return(maxrate * apply(mat, 2, function(vec){
      exp(-t(vec - mu) %*% siginv %*% (vec - mu) / 2)
    }))
}

@@

<<sim2000obs, cache = TRUE, dependson = 'uxoexampledata', fig.pos = 't', fig.cap = 'Observed geomagnetic anomalies from one realization of the UXO site. The transects are not shown on this plot because they would obscure the points.'>>=
# Load the sample data used in my writing project.
# The anomaly file contains the locations of the observed anamolies and the
# cog file contains the course-over-ground waypoints on the transects.
ex_anom <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly')
ex_path <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.cog')

@@

The kernel density estimate is computed using \texttt{spatstat}'s default
``simple rule of thumb'' to select the smoothing bandwidth. Because the
anomalies located more than three feet from a transect were not observed, I
would expect the kernel density estimate to show ridges of high intensity
along the transects. As it turns out, the default bandwith is large enough to
smooth across multiple transects so the ridges do not appear. However, the
kernel density estimate is biased low because not all anomalies are observed;
I divide the estimate by the proportion of the area observed as an approximate
bias correction and the result is a surprisingly reasonable intensity map.

@@

Figure~\ref{fig:polyse1} shows the log-scale standard errors for six of the
polynomial models. For all models except the twelfth-degree model, the
standard error is low across most of the site but high along the boundary. The
standard errors increase with the degree of the model. This is expected and
resonable because locations in the interior of the region have the most nearby
anomalies and dummy points, and predictions from more complex models depend on
more estimated parameters.

@@

The troubles with \texttt{spatstat} largely come from how the package sets up
grids of locations for dummy points and predictions. It finds a bounding box
for the region and places grid cells on a lattice across the bounding box, but
its grids are so coarse that most of my transects do not have cells centered
on them. It attempts to create an \(80 \times 80\) grid of dummy points, but
the only useable points are 160 points on two transects; all others are
discarded~(Figure~\ref{fig:dummy}, left). Attempting to fit models in
\texttt{ppm} using these dummy points results in an error.

@@

The biggest disappoinment of using \texttt{spatstat} to fit these models is
that models with splines cannot be plotted. The \texttt{ppm} functions
estimates the parameters without any apparent trouble, but the \texttt{plot}
and \texttt{predict} functions give errors because they do not know how to
interpret the smooth terms in the model formula.

>>>>

One little-studied application of inhomogeneous point process intensity
estimation is the use of an intensity surface estimated from events in a
subregion to predict the intensity over the entire region of interest. This
procedure would be relevant whenever it is known or suspected that some type of
plant, animal, or other item occurs in the region, and the goal of the
analysis is to map the trend in where these tend to be located rather than
estimate parameters of some process at work across hypothetical replicates of
similar regions. It may be prohibitively expensive or difficult to observe all
events over the entire region, so a sample of subregions is taken. For
example, conservationists want to study the spatial distribution of an
endangered plant across a large region of thick jungle so that they can
establish a preserve where the plant is protected. They cannot search the
entire jungle, so they take a simple random sample of quadrats and record the
locations of the plants in those quadrats. They will then want to fit a model
describing the trends in intensity of these plants extrapolated over the
entire jungle. In this setting, the objective is to map the realized spatial
intensity of the plant over this jungle so that these plants can be protected,
not to estimate parameters of the process that arranges plants of this species
at jungles like this one.

Another situation where it is useful to extrapolate point pattern intensity
outside of the observed region is in mapping subsurface geomagnetic anomalies,
such as the munitions debris found at former military test ranges. This is
frequently done in the early stages of an unexploded ordnance (UXO)
remediation, where the intensity surface of inert munitions fragments is
used to identify the locations of targets so that the search for UXO can be
focused on the sections of the site most likely to contain it. The anomalies
are only observed when detection equipment passes directly over them, but the
project leaders are concerned with finding UXO and do not want to waste
resources finding every inert fragment that could be found by metal detectors.
The most common data collection method is to take a systematic sample of
straight-line transects and observe anomalies in rectangular regions centered
along the transects. Frequently, moving averages of the intensity are computed
in circular windows and an intensity map is produced using ordinary kriging to
predict the intensity in a grid of these windows. There are several problems
with this approach: it assumes stationarity when the moving averages are
believed to be non-stationary, it ignores the point process nature of the
data, and it is very sensitive to the window size~\parencite{vspguide,wp}. In
this project, I review the use of polynomial trend surface models for the
log-intensity at a simulated UXO site, ultimately concluding that these
methods are not yet developed well enough to be useful.

@@

The log-likelihood of an unmarked Poisson process on a region \(D\) with
intensity \(\lambda(\mathbf{s})\) is found by conditioning on the number of
events in the following manner. The number of events in \(D\) follows a
Poisson distribution with mean \(\mu=\int_{D}\lambda(\mathbf{s})d\mathbf{s}\).
Given that \(n\) events occurred, their locations \(\mathbf{s}_{1},\dots,
\mathbf{s}_{n}\) are independent and identically distributed with density
\(\lambda(\mathbf{s})/\mu\). Then the log-likelihood of the intensity is
\begin{align*}
\ell(\lambda) &= \left\{-\mu + n\log(\mu) - \log(n!)\right\}
+ \sum_{i=1}^{n}\left\{\log\left(\lambda(\mathbf{s}_{i})\right)
- \log(\mu)\right\} \\
&= \sum_{i=1}^{n}\log\left(\lambda(\mathbf{s}_{i})\right)
-\int_{D}\lambda(\mathbf{s})d\mathbf{s} - \log(n!).
\end{align*}
A natural way to model the intensity is to define a log-linear model
\begin{equation*}
\log(\lambda(\mathbf{s})) = \mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
\end{equation*}
where \(\mathbf{x}(\mathbf{s})\) can, in principle, include functions of the
spatial coordinates and also covariates. Unfortunately, the likelihood becomes
\begin{equation*}
\sum_{i=1}^{n}\mathbf{x}(\mathbf{s})^{T}\boldsymbol{\beta}
-\int_{D}\exp\left(\mathbf{x}(\mathbf{s})^{T}
\boldsymbol{\beta}\right)d\mathbf{s} - \log(n!)
\end{equation*}
so \(\mathbf{x}(\mathbf{s})\) must be known (or predicted) either across the
entire region \(D\) or at enough locations to approximate the integral
numerically. Thus, covariates require some extra data collection or modeling
effort to incorporate, but trend surfaces are feasible.

\textcite{bermanturner} proposed a practical method for fitting these models
with generalized linear model (GLM) software using quadratic approximations to
the likelihood, even generalizing to other link functions besides the log link.
Their method is based on partitioning \(D\) such that each subset in the
partition contains at most one event, and then maximizing a weighted
pseudolikelihood based upon a binomial or Poisson distribution for the count
of events in each subset. They comment that the covariance matrices produced
by standard GLM software under this method are reasonable approximations to
the correct covariance matrices that would be derived from the true likelihood
as long as the quadrature approximation of the likelihood converges to the
true likelihood as the resolution of the partition is increased.
This was followed up by \textcite{baddeleycoeurjolly14}, who developed a
logistic regression-based approach that avoids the quadratic approximation,
thereby reducing bias in the coefficient estimates. Both methods are available
in the \texttt{spatstat} package~\parencite{spatstat} for R~\parencite{baser}.
The \texttt{spatstat} function \texttt{ppm} implements these methods using the
base R \texttt{glm} function to maximize the likelihood; I will use the Berman
and Turner method because it is the default in \texttt{ppm} and therefore
better reflects the experience of a na\"{i}ve practitioner who is not an
expert on spatial point processes.

@@

In my writing project, I evaluated ordinary kriging of moving average
intensity values on a simulated tank training range. I will use the same site
for this project. The site is a
\Sexpr{sprintf('%.2f', area(site_window) / 43560)} acre quadrilateral region
with homogeneous Poisson process producing background anomalies with an
intensity of 100 anomalies per acre, and two regions of concentrated munitions
use generated as Poisson processes with bivariate Gaussian intensity. The
total intensity at the center of each concentrated munitions use region is
300 anomalies per acre~(Figure~\ref{fig:truth}).

<<truth, cache = TRUE, dependson = 'uxoexampledata', fig.width = 4, fig.pos = 't', fig.cap = 'True intensity at the simulated UXO site.'>>=
# Bivariate normal intensity function that I originally wrote for my
# writing project, parameterized in terms of the major and minor axes scales
# and the angle of the major axis from horizontal.
#   a is horizontal axis, b is vertical axis, r is rotation angle
gauss.elliptic <- function(x, y, mu.x = 0, mu.y = 0, s.a = 1, s.b = 1,
                           r = 0, maxrate = 1){
  rot <- zapsmall(matrix(c(cos(r), sin(r), -sin(r), cos(r)), nrow = 2))
  ab <- diag(c(s.a^2, s.b^2))
  sigma <- rot %*% ab %*% t(rot)
  siginv <- solve(sigma)
  mu <- matrix(c(mu.x, mu.y), nrow = 2)
  mat <- matrix(rbind(x, y), nrow = 2)
  return(maxrate * apply(mat, 2, function(vec){
      exp(-t(vec - mu) %*% siginv %*% (vec - mu) / 2)
    }))
}

@@

<<sim2000obs, cache = TRUE, dependson = 'uxoexampledata', fig.pos = 't', fig.cap = 'Observed geomagnetic anomalies from one realization of the UXO site. The transects are not shown on this plot because they would obscure the points.'>>=
# Load the sample data used in my writing project.
# The anomaly file contains the locations of the observed anomalies and the
# cog file contains the course-over-ground waypoints on the transects.
ex_anom <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.anomaly')
ex_path <- read.csv('easy_sample_tTA2_p200_bg100_fg200_rep2000.cog')

@@

The kernel density estimate is computed using \texttt{spatstat}'s default
``simple rule of thumb'' to select the smoothing bandwidth. Because the
anomalies located more than three feet from a transect were not observed, I
would expect the kernel density estimate to show ridges of high intensity
along the transects. As it turns out, the default bandwidth is large enough to
smooth across multiple transects so the ridges do not appear. However, the
kernel density estimate is biased low because not all anomalies are observed;
I divide the estimate by the proportion of the area observed as an approximate
bias correction and the result is a surprisingly reasonable intensity map.

@@

Figure~\ref{fig:polyse1} shows the log-scale standard errors for six of the
polynomial models. For all models except the twelfth-degree model, the
standard error is low across most of the site but high along the boundary. The
standard errors increase with the degree of the model. This is expected and
reasonable because locations in the interior of the region have the most nearby
anomalies and dummy points, and predictions from more complex models depend on
more estimated parameters.

@@

The troubles with \texttt{spatstat} largely come from how the package sets up
grids of locations for dummy points and predictions. It finds a bounding box
for the region and places grid cells on a lattice across the bounding box, but
its grids are so coarse that most of my transects do not have cells centered
on them. It attempts to create an \(80 \times 80\) grid of dummy points, but
the only usable points are 160 points on two transects; all others are
discarded~(Figure~\ref{fig:dummy}, left). Attempting to fit models in
\texttt{ppm} using these dummy points results in an error.

@@

The biggest disappointment of using \texttt{spatstat} to fit these models is
that models with splines cannot be plotted. The \texttt{ppm} functions
estimates the parameters without any apparent trouble, but the \texttt{plot}
and \texttt{predict} functions give errors because they do not know how to
interpret the smooth terms in the model formula.
